Mathematical Language Processing \\ Project

### 2014-06-15 paper.tex
current state of the article

\usepackage{pgfplots}
\usepackage{tikz}
\usetikzlibrary{arrows,backgrounds,positioning}

@@

In natural language, words and phrases themselves imply the semantics. In
contrast, the meaning of identifiers in mathematical formulae is undefined.
Thus scientists must study the context to decode the meaning. The
\emph{Mathematical Language Processing} project aims to support that process
and presents automated identifier definition discovery. Ordinary natural
language processing approaches are trained on large, manually annotated text
corpora. They fall short in reasonable quality of definition discovery for
scientific texts. We use statistical methods to discover definitions in the
surrounding text. The evaluation of our prototypical system, applied on the
Wikipedia text corpus, shows that our approach augments the user experience
substantially. While hovering the identifiers in the formula, tool-tips with
the most probable definitions occur. Tests with random samples show that the
displayed definitions provide a good match with the actual meaning of the
identifiers.

@@

@@

Extracting relevant identifiers from the article rely on the assumption that
the author will use \texttt{<math/>} tags for all formulae. Thus said, a
formula that is written within a text cannot be recognized, and therefor, any
identifier that is solely used there, can never be extracted.

The fact that we can estimate all relevant identifiers for an article (see
Section \ref{ir}), combined with some common assumptions about the relation,
can be exploited to largely reduced set of candidates that need to be ranked.
Please note that this reduction is essential for retrieving the correct
relations. Otherwise almost any word would be ranked and the precision of the
retrieval would drop significantly.

The basic assumption of our approach is that if two entities take part in a
relation, those two co-occur in the same sentence. Any other sentence is,
therefor, irrelevant. Another assumption can be made about the lexical class
of the desciption we want to rank. They are nouns or even noun phrases (e.g.
\texttt{ `the effective spring constant k' } or \texttt{ `mass m of
something'} ). Beside discarding all other words (according to their POS tag),
noun phrases and terms within a Wikipedia hyperlink will be aggregated into
one new candidate term. This is important, due to the fact that overall
ranking will be greatly infuenced by the distance of candidate words. For all
intends and purposes, it is not needed to keep noun phrases as single words.
So they can be aggregated and be ranked as if they were a single word.

@@

\subsubsection{Identifier Retrieval}
\label{ir}

@@

We took a very conservative approach and preprocessed all formulas.
\texttt{\textbackslash text\{\}} blocks along with subscriptions containing
more than a single character will be removed before analysis. Superscripts
will also be ignored in terms of being a part of the identifier. Moreover, we
created a comprehensive blacklist to improve the results further. Short words
like `a' and `i', which are very common in the english language, could be
easily matched by our processor in the sourrunding text if a formula contains
one of those identifier, and therefor, will also be blacklisted. Additionally,
we blacklist common mathematical operators, constants, and functions.

@@

\subsection{Description Retrieval}

We ran the program on a dataset of about 20.000 articles, all containing
\texttt{<math/>} tags, and retrieved about 550.000 candidate relations. The
most common definition relations were:

\begin{tabular}{| l | p{6.8cm} | l |}
	\hline
	Identifier & Descriptions & Count\\
	\hline
	$n$ & number & 1709 \\
	$t$ & time & 1480 \\
	$M$ & mass & 1042 \\
	$r$ & radius & 752 \\
	$T$ & temperature & 666 \\
	$\theta$ & angle & 639 \\
	$G$ & group & 635 \\
	\hline
\end{tabular}

\paragraph{Observations.} We observed some poorly ranked relations. For
example, in the fragment \texttt{`where $\phi$ ( $r_{i}$ ) is the electrostatic
potential'} the distance
$\Delta(\phi, \mathsf{electrostatic potential} ) = 5$. This is due
to counting brackets and function arguments as words. Also wrongly tagged
words like `Hamiltonian/JJ' lead to false negatives.

@@

As there were no evaluation datasets available to us, we compared our
statistical ranking to a simple pattern matching approach. We used the
following patterns to discover description terms:

\begin{tabular}{| p{9.3cm} |}
	\hline
	Pattern \\
	\hline
	\texttt{\emph{<description>} \emph{<identifier>}} \\
	\texttt{\emph{<identifier>} is \emph{<description>}} \\
	\texttt{\emph{<identifier>} is the \emph{<description>}} \\
	\texttt{let \emph{<identifier>} be the \emph{<description>}} \\
	\texttt{\emph{<description>} is|are denoted by \emph{<identifier>}} \\
	\texttt{\emph{<identifier>} denotes */DT \emph{<identifier>}} \\
	\hline
\end{tabular}

\paragraph{} Due to large amount of time needed to evaluate two approaches by hand on a
multitude of articles, we build a dataset of definition relations (98
relations in total), that can be retrieved from two representative articles
(revision ids are included). This dataset is also available on the project
repository.

\begin{tabular}{| l | l | l | l |}
	\hline
	 & MLP-Ranking ($k=1$) & MLP-Ranking ($k=2$) & Pattern Matching \\
	\hline
	Precision &  0.872  &  0.915  &  0.911  \\
	Recall    &  0.839  &  0.892  &  0.733  \\
	\hline
\end{tabular}

Note: $k$ equals the amount of the top ranked candidate definitions.

% \paragraph{} We than took the pattern matching approach as a baseline for our ranked
% retrievals sets as these patterns cover most common definition relations we
% found so far.

@@

The distance measure $R_{\sigma_d}$ fails for the example of
Fig.~\ref{fig:screenshot} since $\Delta(\mathsf{energy},E) =
\Delta(\mathsf{mass},m) = 2$. Unfortunatly one cannot simply detect
punctuation marks and introduce some kind of directed associativity (e.g.
inflicting a penatly on the ranking if the relations spans over a comma). That
would lead to whole classes of relation `types' (in terms of the grammatical
structure) never being retrieved. We plan to mitigate this problem by taking
more closely related scientific articles (based on their specific fields) into
consideration and count the frequencies of the candidate relations. The
intuition behind this is, that articles of the same scientific field will
likely use the same definition for the identifiers.

>>>>

\usepackage{pgfplots}
\usepackage{tikz}
\usetikzlibrary{arrows,backgrounds,positioning}
\usepackage{todonotes}

@@

In natural language, words and phrases themselves imply the semantics. In
contrast, the meaning of identifiers in mathematical formulae is undefined.
Thus scientists must study the context to decode the meaning. The
\emph{Mathematical Language Processing} project aims to support that process
and presents automated identifier definition discovery. We compare two
approaches. At first we use a simple identifier definition pattern matching
approach. Second, we present the \emph{MLP} approach, that uses POS tag based
distances as well as sentence positions to caculate indentifier definition
probabilities. The evaluation of our prototypical system, applied on the
Wikipedia text corpus, shows that our approach augments the user experience
substantially. While hovering the identifiers in the formula, tool-tips with
the most probable definitions occur. Tests with random samples show that the
displayed definitions provide a good match with the actual meaning of the
identifiers.

@@

\section{Pattern-based Definition Discovery}

At first, we implemented a simple identifier definition extractor that is
based on a set of static patterns. As it is an fairly robust approach und easy
to implement, this will serve us as a good reference point in terms of
performance. The following patterns are being used to discover description
terms:

\begin{table}
\vspace{-5pt}
	\begin{center}
		\begin{tabular}{| p{9.3cm} |}
			\hline
			Pattern \\
			\hline
			\texttt{\emph{<description>} \emph{<identifier>}} \\
			\texttt{\emph{<identifier>} is \emph{<description>}} \\
			\texttt{\emph{<identifier>} is the \emph{<description>}} \\
			\texttt{let \emph{<identifier>} be the \emph{<description>}} \\
			\texttt{\emph{<description>} is|are denoted by \emph{<identifier>}} \\
			\texttt{\emph{<identifier>} denotes */DT \emph{<description>}} \\
			\hline
		\end{tabular}
	\end{center}
\caption{Implemented static patterns}
\vspace{-5pt}
\end{table}

Note that in the last pattern we used a \emph{part-of-speech} tag. Actually
determiners not only contain articles but also quatifiers and distributives.
To increase readability this list have been shortened to `*/DT'.

@@

Extracting relevant identifiers from the article rely on the assumption that
the author will use \texttt{<math/>} tags for all formulae. This said, a
formula that is written within a text cannot be recognized, and therefor, any
identifier that is solely used there, can never be extracted.

The fact that we can estimate all relevant identifiers for an article (see
Section \ref{ir}), combined with some common assumptions about the relation,
can be exploited to largely reduce the set of candidates that need to be ranked.
Please note that this reduction is essential for retrieving the correct
relations. Otherwise almost any word would be ranked and the precision of the
retrieval would drop significantly.

The basic assumption of outr approach is that the two entities of a definition
relation co-occur in the same sentence.

The basic assumption of our approach is that if two entities take part in a
relation, those two co-occur in the same sentence.
Any other sentence is, therefor, irrelevant.

Another assumption can be made about the lexical class
of the desciption we want to rank. They are nouns or even noun phrases (e.g.
\texttt{ `the effective spring constant k' } or \texttt{ `mass m of
something'} ). Beside discarding all other words (according to their POS tag),
noun phrases and terms within a Wikipedia hyperlink will be aggregated into
one new candidate term. This is important, due to the fact that overall
ranking will be greatly infuenced by the distance of candidate words. For all
intends and purposes, it is not needed to keep noun phrases as single words.
So they can be aggregated and be ranked as if they were a single word.

@@

\subsection{Identifier Retrieval}
\label{ir}

@@

We took a very conservative approach and preprocessed all formulas.
\texttt{\textbackslash text\{\}} blocks along with subscriptions containing
more than a single character will be removed before analysis. Superscripts
will also be ignored in terms of being a part of the identifier. Moreover, we
created a comprehensive blacklist to improve the results further. Short words
like `a' and `i', which are very common in the English language, could be
easily matched by our processor in the surrounding text if a formula contains
one of those identifier, and therefore, will also be blacklisted. Additionally,
we blacklist common mathematical operators, constants, and functions.

@@

\subsection{Description Retrieval}
We ran the program on a dataset of about 20.000 articles, all containing
\texttt{<math/>} tags, and retrieved about 550.000 candidate relations. The
most common definition relations were:

\begin{table}[H]
	\begin{center}
	\begin{tabular}{| l | p{6.8cm} | l |}
		\hline
		Identifier & Descriptions & Count\\
		\hline
		$n$ & number & 1709 \\
		$t$ & time & 1480 \\
		$M$ & mass & 1042 \\
		$r$ & radius & 752 \\
		$T$ & temperature & 666 \\
		$\theta$ & angle & 639 \\
		$G$ & group & 635 \\
		\hline
	\end{tabular}
	\end{center}
\caption{Most common definition relations}
\end{table}

\paragraph{Observations.} We observed some poorly ranked relations. For
example, in the fragment \texttt{`where $\phi$ ( $r_{i}$ ) is the electrostatic
potential'} the distance
$\Delta(\phi, \mathsf{electrostatic\:potential} ) = 6$. This is due
to counting brackets and function arguments as words. Also wrongly tagged
words like `Hamiltonian' as an adjective leads to false negatives. Latter

@@

Due to the lack of gold standard evaluation datasets to measure the
performance of identifer definition extractors, we've created on on our own.
This is a very time consuming job. At the moment the dataset only contains two
large articles (revision ids are included) with around 100 identifier
definitions. This dataset is also available on the project repository.

As in lots of articles, the ones in the evaluation dataset contain identifiers
whose description cannot be retrieved. This is due to two reasons. First and
foremost, the identifier found in a formula is never mentioned in the
surrounding text, and therefor, no description can be extracted. Second, the
identifier is somehow ambigious (see section~\ref{ir}) and has been dropped. Most
notably, identifiers like $I_{xx}$ will be discarded because of an ambigious
index that contains multiple letters.

Unfortunately 32 out of 99 identifiers from our dataset fall into that category.
We've decided to evaluate the performance of the remainder, as those 32 do not
convey any conceptual flaws. From the users standpoint, the overall performance
(in terms of recall) of such a system whould be rather annoying. As we are only
interested in evaluating the performance of the \emph{MLP Ranking} algoritm itself,
it is save to ignore those 32 identifers.

\begin{table}[H]
\vspace{-5pt}
	\begin{center}
		\begin{tabular}{| l | l | l | l |}
			\hline
			 & MLP-Ranking ($k=1$) & MLP-Ranking ($k=2$) & Pattern Matching \\
			\hline
			Precision &  0.872  &  0.915  &  0.911  \\
			Recall    &  0.839  &  0.892  &  0.733  \\
			\hline
		\end{tabular}
	\end{center}
\caption{Evaluation results. Note: $k$ equals the amount of the top ranked candidate definitions.}
\vspace{-20pt}
\end{table}

The results show that the unoptimized MLP approach keeps up with the
performance of the simple pattern matcher. Furthermore, we observed that it is
more robust in terms of recall, as it is less vulnerable to small changes in
the sentence structure.

@@

The distance measure $R_{\sigma_d}$ fails for the example of
Fig.~\ref{fig:screenshot} since $\Delta(\mathsf{energy},E) =
\Delta(\mathsf{mass},m) = 2$. Unfortunately, one cannot simply detect
punctuation marks and introduce some kind of directed associativity (e.g.
inflicting a penalty on the ranking if the candidate relations spans over a
comma). That would lead to whole classes of relation `types' (in terms of the
grammatical structure) never being retrieved. We plan to mitigate this problem
by taking more closely related scientific articles (based on their specific
fields) into consideration and count the frequencies of the candidate
relations. The intuition behind this is, that articles of the same scientific
field will likely use the same definition for the identifiers. Moreover,
we hope to also mitigate the problem of `dangling' identifiers (those not
mentioned in the article itself), as their may be description in related
articles.
